Stable Hot Spot Analysis (Draft)

### 2017-06-12 paper-main.tex

% ==========================================
% Support for \FOCAL{R}{M}{op} (by Viliam)
% ==========================================
\newcommand{\FOCALMMODE}[3]{
  #1{\stackrel{\mathtt{#3}}{\circ}}#2
}
\newcommand{\FOCAL}[3]{
  \ifmmode
  \FOCALMMODE{#1}{#2}{#3}
  \else
  \begin{math}\FOCALMMODE{#1}{#2}{#3}\end{math}%
  \fi
}
% ==========================================

@@

Existing methods to determine hot spots are dependent on the parametrization of
the weight matrix as well as on the size of the study area.

Consider the real-world example depicted in Fig.~\ref{fig:TempMaps}. The
temperature map of a morning thermal flight dataset (Fig.~\ref{fig:TempMaps:a})
has been processed using $G^*$.

\begin{figure}[htp]
  \begin{subfigure}{1\linewidth}
    \caption{Morning temperatures}
    \includegraphics[width=\linewidth]{images/hotspot-rawtemp-1}
    \label{fig:TempMaps:a}
  \end{subfigure}

  \caption{
    Karlsruhe city center.
    Selected area of $2.4{\times}1.4\,km$.
    Pixel size $5{\times}5\,m$.
  }
  \label{fig:TempMaps}
\end{figure}

% Bild besteht aus orginal, weight matrix a und weight matrix b
As we can see, hot spots are oftentimes disappearing or appearing unrelated to
previously found hot spots. While these computations indeed show hot spots and
the results are correct, they lack stability.

For a data analyst, when exploring the data interactively by choosing different
filter sizes (in form matrices), it is important that the hot spot's position
and size changes in a predictable manner. This intuition is the basis for our
stability metric.

We define a hot spot found in comparably more coarse resolutions as
\emph{parent} (larger weight matrix) and in finer resolutions as \emph{child}
(smaller weight matrix). To be stable, one assumes that every parent has at
least one child and that each child has one parent. For a perfectly stable
interaction, it can be easily seen that the connection between parent and child
is a injective function and between child and parent a surjective function. To
measure the closeness of connection, we propose a metric called the
\emph{Stability of Hot spot} (SoH). It measures the deviation from a perfectly
stable transformation of resolutions.

%TODO: what about the other direction?
In its downward property (from parent to child, injective) it is defined as:
\begin{equation}
\label{eq:SoH-down}
SoH^\downarrow = \frac{ParentsWithChildNodes}{Parents} = \frac{|Parents \cap 
Children|}{|Parents|}
\end{equation}
And for its upward property (from child to parent, surjective):
\begin{equation}
\label{eq:SoH-up}
SoH^\uparrow = \frac{ChildrenWithParent}{Children} = 1 - \frac{|Children - 
Parents|}{|Children|}
\end{equation}
where $ParentsWithChildNodes$ is the number of parents that have at least one 
\emph{child}, $Parents$ is the total number of \emph{parent}, 
$ChildrenWithParent$ is the number of children and $Children$ as the total 
number of children.
The SoH is defined for a range between 0 and 1, where 1 represents a perfectly 
stable transformation while 0 would be a transformation with no stability at 
all.

@@

@@

>>>>

% ==========================================
% Custom commands (by Viliam)
% ==========================================
\newcommand{\FOCAL}[3]{ % usage: \FOCAL{R}{M}{op}
  \ifmmode
    #1{\stackrel{\mathtt{#3}}{\circ}}#2
  \else
    \begin{math}\FOCAL{#1}{#2}{#3}\end{math}
  \fi
}
\newcommand{\SOHUP}{
  \ifmmode
    SoH^\uparrow
  \else
  \begin{math}\SOHUP\end{math}
  \fi
}
\newcommand{\SOHDOWN}{
  \ifmmode
    SoH^\downarrow
  \else
  \begin{math}\SOHDOWN\end{math}
  \fi
}
% ==========================================

@@

Existing methods for determining hot spots are dependent on the parametrization
of the weight matrix as well as on the size of the study area. Intuitively,
increasing the size of a weight matrix has a "blurring" effect on the raster
(Fig.~\ref{def:TODO}a) whereas decreasing the size can be seen as a form of
"sharpening" (Fig.~\ref{def:TODO}b).

%TODO: add image showing and blurring/sharpening effects

%As we can see, hot spots are oftentimes disappearing or appearing unrelated to
%previously found hot spots. While these computations indeed show hot spots and
%the results are correct, they lack stability.

For a data analyst, when exploring the data interactively by choosing different
filter sizes (weight matrices) or point aggregation strategies (pixel sizes), it
is important that the position and size of a hotspot changes in a predictable
manner. We formalize the intuition in our stability metric.

%
%Consider the real-world example depicted in Fig.~\ref{fig:TempMaps}.
%%TODO: thermal flight -> NY taxi
%The temperature map of a morning thermal flight dataset 
%(Fig.~\ref{fig:TempMaps:a}) has been processed using $G^*$.
%
%%TODO: thermal flight -> NY taxi
%%TODO: need two images - G* and FocalG*
%\begin{figure}[htp]
%  \begin{subfigure}{1\linewidth}
%    \caption{Morning temperatures}
%    \includegraphics[width=\linewidth]{images/hotspot-rawtemp-1}
%    \label{fig:TempMaps:a}
%  \end{subfigure}
%  
%  \caption{
%    Karlsruhe city center.
%    Selected area of $2.4{\times}1.4\,km$.
%    Pixel size $5{\times}5\,m$.
%  }
%  \label{fig:TempMaps}
%\end{figure}

We define a hot spot found in comparably more coarse resolutions as
\emph{parent} (larger weight matrix or larger pixel size) and in finer
resolutions as \emph{child} (smaller weight matrix or smaller pixel size).

That is

To be stable, one assumes that every parent has at
least one child and that each child has one parent. For a perfectly stable
interaction, it can be easily seen that the connection between parent and child
is a injective function and between child and parent a surjective function. To
measure the closeness of connection, we propose a metric called the
\emph{Stability of Hot spot} (SoH). It measures the deviation from a perfectly
stable transformation of resolutions.

In its downward property (from parent to child, injective) it is defined as:
\begin{equation}
  \label{eq:SoH-down}
  \SOHDOWN
    = \frac{ParentsWithChildNodes}{Parents}
    = \frac{|Parents \cap Children|}{|Parents|}
\end{equation}
And for its upward property (from child to parent, surjective):
\begin{equation}
  \label{eq:SoH-up}
  \SOHUP
    = \frac{ChildrenWithParent}{Children}
    = 1 - \frac{|Children - Parents|}{|Children|}
\end{equation}
where $ParentsWithChildNodes$ is the number of parents that have at least one 
\emph{child}, $Parents$ is the total number of \emph{parent}, 
$ChildrenWithParent$ is the number of children and $Children$ as the total 
number of children.
The SoH is defined for a range between 0 and 1, where 1 represents a perfectly 
stable transformation while 0 would be a transformation with no stability at 
all.

@@

\section{Evaluation}

- To evaluate, we compare G* with FocalG* on the same dataset.
- We use NY taxi dropoffs
- a single evaluation run is defined in Def.~\ref{def:EvalRun}.
- In an ideal case, we could produce a 3D-plot:
  - x-axis would be growing pixel sizes 100..1000 by 100 (aka zoom-out)
  - y-axis would be growing weight matrix sizes (e.g. 5..41 by 2)
  - z-axis would be performance of G* vs FocalG* x and y coordinates
  - we could also repeat the computation for multiple datasets to obtain 
  multiple samples and to compute error bars.
- due to paper size limit, we choose projection in which matrix size is set to 
constant 7x7 pixels and the only variable is the pixel size. This way, we 
produce a 2D plot. We also generate just a single sample.

\begin{definition}[Evaluation Run]
  \label{def:EvalRun}
  We define a single evaluation run as a tuple:
  \begin{displaymath}
    E = (V, m, p, w)
  \end{displaymath}
  \noindent where:
  \begin{itemize}

    \item $V$ is the input dataset of points, representing the taxi dropoffs in 
    our case.

    \item $m$ is the metric used, in our case either \SOHUP or \SOHDOWN.

    \item $p$ represents the pixel size for aggregating points from $V$, 
    e.g. $100 \times 100$ meters.

    \item $w$ represents the size of a weight matrix.
    In our case, we chose a weight matrix depicted in 
    Figure~\ref{fig:ExampleMatrices}(a) for both the G* and FocalG* cases.
  \end{itemize}
\end{definition}

%Fig.~\ref{fig:ComparisonMorning} and Fig.~\ref{fig:ComparisonEvening} show
%Standard and Focal $G^*$ computations for both morning and eventing datasets
%with weight matrix $W$ of size 3, 5, 7, 9, 15 and 31. In this figures, when 
%computing Focal
%$G^*$, the focal matrix $F$ has a constant size of $61{\times}61$ cells. 
%Example weight matrix $W$ and focal matrix $F$ are depicted in 
%Fig.~\ref{fig:ExampleMatrices}.

@@

The evaluation results are plotted in Fig.~\ref{fig:EvalResults}, each point in
the graph represents the \SOHUP metric (Eq.\ref{eq:SoH-up}) between two
$G^*$ generated using weight matrices of size $i$ and $i+2$. The focal matrix F
has a fixed size of $41{\times}41$

\begin{displaymath}
\SOHUP( G^*(R, W_i, st), G^*(R, W_{i+2}, st) )
\end{displaymath}
